Claret: Using Data Types for Highly Concurrent Distributed Transactions

### 2015-02-14 rmd/paper.Rmd
minor wording changes, fix urls in bibliography

In situations where consistency really is crucial, developers can enforce stronger guarantees on top of any platform, but they often get it wrong. Google, recognizing the benefit to programmer productivity, reintroduced serializable transactions to their next generation database, Spanner \cite{Spanner}. However, this leaves them with two extreme choices, with significant performance impact. If certain parts of an application can tolerate imprecision, then why not capture those properties in the database programming model? What if the programmer could express the semantics they desire succinctly and precisely, in a way that the database can better reason about conflicts and optimize performance, without giving up the flexibility and scalability of their existing systems?

Abstract data types (ADTs) are the solution to this problem. Rather than limiting the records in databases to being primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers. Understanding the properties of ADT operations, such as which operations commute, can be used in a variety of ways to improve database performance, such as reducing the number of aborts. By choosing the right data type, programmers make tradeoffs between performance and precision based on their application's needs.

@@

Though *commutativity* is often discussed in terms of an operation commuting with all other operations, it is actually more nuanced. If a pair of operations commute, then executing them in either order will produce the same result. Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a function of the methods, their arguments, their return values, and the *abstract* state of their target. We call the full set of commutativity rules for an ADT its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}. It is important to note that interface choice affects commutativity: if `add(x)` returned a boolean that expressed if the item was added or not, then `add(x)` would only commute with `add(y)` if $x \ne y$. In this case, the difference doesn't drastically impact commutativity, but it would affect the cost of checking commutativity dynamically. There are actually many valid specifications which expose less than the maximum commutativity, but may be cheaper to implement.\cite{Kulkarni:PLDI11} ADT designers can navigate these choices to optimize an implementation.

## Transactional Boosting
If two operations on the same record in two different transactions commute, then the transactions can safely execute concurrently, even though they both update the record. This technique of raising the level of abstraction in transactions to operations is known as *transactional boosting* \cite{Herlihy:PPoPP08}. This straightforward use of commutativity was shown to significantly improve abort rates in state-of-the-art software transactional memory. In \autoref{eval}, we show how it can be applied to distributed transactions.

@@

@@

\autoref{tput} shows the results of this simulation. When most of the traffic is content consumption (reading timelines), both systems perform well enough. However, when we simulate a workload where clients repost popular posts from their timelines, we see a viral propagation effect, where a large fraction of the users get and share a post. As Twitter came to a standstill when Ellen Degeneres's Oscar selfie set a retweeting record, so too does our baseline fall over. But with commutativity, we see that performance continues to scale even under this arduous load.

>>>>

In situations where consistency really is crucial, developers can enforce stronger guarantees on top of any platform, but they often get it wrong. Recognizing the benefit to correctness and productivity, some systems have reintroduced serializable transactions, such as Google's Spanner \cite{Spanner}. However, this leaves programmers with two extreme choices with significant performance impact. If certain parts of an application can tolerate imprecision, then why not capture those properties in the database programming model? What if the programmer could express the semantics they desire succinctly and precisely, in a way that the database can better reason about conflicts and optimize performance, without giving up the flexibility and scalability of their existing systems?

Abstract data types (ADTs) are the solution. Rather than limiting the records in databases to being primitive types like strings or integers, raising them to more complex data types provides a richer interface, exposing ample opportunities for optimization to the database and a precise mechanism to express the intentions of programmers. Understanding the properties of ADT operations, such as which operations commute, can be used in a variety of ways to improve database performance, such as reducing the number of aborts. By choosing the right data type, programmers make tradeoffs between performance and precision based on their application's needs.

@@

Though *commutativity* is often discussed in terms of an operation commuting with all other operations, it is actually more nuanced. If a pair of operations commute, then executing them in either order will produce the same result. Using the definitions from \cite{Kulkarni:PLDI11}, whether or not a pair of method invocations commute is a function of the methods, their arguments, their return values, and the *abstract* state of their target. We call the full set of commutativity rules for an ADT its *commutativity specification.* An example specification for a *Set* is shown in \autoref{spec}. There are actually many valid specifications which expose less than the maximum commutativity, but may be cheaper to implement.\cite{Kulkarni:PLDI11}

## Transactional Boosting
If two operations on the same record in two different transactions commute, then the transactions can safely execute concurrently, even though they both update the record. This technique is known as *transactional boosting* \cite{Herlihy:PPoPP08}. This straightforward use of commutativity was shown to significantly improve abort rates in software transactional memory. In \autoref{eval}, we show how we applied it to distributed transactions.

@@

# Data type selection

Choosing the right data type for your application can drastically impact performance.
Selecting the semantics that are most permissive or specialized for a particular use case gives the system the best chance of scaling performance.
For a simple example, an application needing to generate unique IDs should not use a counter, which must return the next number in the sequence, because this is very difficult to scale (as users of TPC-C \cite{TPCC}, which explicitly requires this, know well). Instead, a `UniqueID` type succinctly expresses that non-sequential IDs are okay, which can be implemented very efficiently.
By allowing approximations or non-determinism, performance may be further improved.

**Probabilistic data types** are a class of data types with *probabilistic* guarantees about their semantics, trading off some accuracy for better performance or storage. These include *bloom filters* \cite{bloom}, *hyperloglog* \cite{hyperloglog}, and *count-min sketch* \cite{countminsketch}. Hyperloglog, which also appears in Redis \cite{redis}, estimates the size of a set within a fixed error bound. Twitter's streaming analytics system \cite{summingbird} leverages these to handle their high data volume, and we can expect similar improvements.

**Conflict-free replicated data types (CRDTs)**, which were invented for eventual consistency, can actually be fit into our model as well, as a new kind of data type. This could allow copies to exist in different shards, asynchronously updating each other. By defining the same kind of *merge* function as traditional CRDTs, these replicas could ensure they all converge to the same state. Clients of these might find them more difficult to reason about, they might making this tradeoff in parts of the application where it otherwise cannot scale.

@@

\autoref{tput} shows the results of this simulation. When most of the traffic is content consumption (reading timelines), both systems perform well enough. However, when we simulate a workload where clients repost popular posts from their timelines, we see a viral propagation effect, where a large fraction of the users get and share a post. As Twitter came to a standstill when Ellen Degeneres's Oscar selfie set a retweeting record, so too does our baseline fall over. But with commutativity, performance continues to scale even under this highly contentious load.
